---
title: "Description of the dataset"
description: |
  Core features of the data, variable transformations, and data splitting process.
author:
  - name: Jim - Claire - Thuy
    url: https://www.uoregon.edu/
date: 12-04-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidymodels)
library(rio)
library(reactable)

theme_set(theme_minimal(15) +
            theme(legend.position = "bottom",
                  panel.grid.major.x = element_line(color = "gray60"),
                  panel.grid.minor.x = element_blank(),
                  panel.grid.major.y = element_blank())
          )
```

# Description of the data 

The original `training` dataset included 39 variables. We randomly sampled 7% of this dataset to reduce the total to 13,260 observations. To expand the number of predictors in our models, the following steps were taken: 

  1. `FRL` (6 variables) and `student_count` (2 variables) imported and joined together by `ncessch` variable.
  2. Two variables created in the `FRL` dataset: 
  
      + proportion of students who qualify for free and reduced lunch 
      + proportion of students who qualify for reduced price lunch
    
  3. `Ethnicities` data imported 
  4. The original 7% sample of the `training` data was joined with the `FRL` and `Ethnicities` datasets to increase the total number of variables to 55 across `3,263 observations. 

```{r, include=FALSE}
# load the given data
set.seed(3000)
data <- read_csv(here::here("data", "train.csv")) %>% 
  select(-classification)

data <- dplyr::sample_frac(data, size = 0.07) #trying 7% 

# get freelunch data

frl <- import("https://nces.ed.gov/ccd/Data/zip/ccd_sch_033_1718_l_1a_083118.zip",
              setclass = "tbl_df")  %>% 
  janitor::clean_names()  %>% 
  filter(st == "OR")  %>%
  select(ncessch, lunch_program, student_count)  %>% 
  mutate(student_count = replace_na(student_count, 0))  %>% 
  pivot_wider(names_from = lunch_program,
              values_from = student_count)  %>% 
  janitor::clean_names()  %>% 
  mutate(ncessch = as.double(ncessch))

stu_counts <- import("https://github.com/datalorax/ach-gap-variability/raw/master/data/achievement-gaps-geocoded.csv",
                     setclass = "tbl_df")  %>% 
  filter(state == "OR" & year == 1718)  %>% 
  count(ncessch, wt = n)  %>% 
  mutate(ncessch = as.double(ncessch))

frl <- left_join(frl, stu_counts)

frl <- frl %>% 
 mutate(prop_frl = free_lunch_qualified / n,
       prop_reduced_lunch = reduced_price_lunch_qualified / n)

# ethnicities

sheets <- readxl::excel_sheets(here::here("data",
"fallmembershipreport_20192020.xlsx"))

ode_schools <- readxl::read_xlsx(here::here("data",
"fallmembershipreport_20192020.xlsx"), sheet = sheets[4])

ethnicities <- ode_schools %>%
select(attnd_schl_inst_id = `Attending School ID`,
sch_name = `School Name`,
contains("%")) %>%
janitor::clean_names()
names(ethnicities) <- gsub("x2019_20_percent", "p", names(ethnicities))

# join
data <- left_join(data, ethnicities)
data <- left_join(data, frl)
```

## Score Variable {.tabset .tabset-fade .tabset-pills}

We're interested in predicting the students' score based on their attributes. The range of the outcome variable is as follows:

### Summary

```{r}
summary(data$score)
```

### Distribution

The distribution of of the 7% random sample of scores is a normal distribution. Additionally, there are no apparent skew or outliers. 

```{r, include=TRUE}
ggplot(data, aes(score)) +
  geom_histogram(fill = "#56B4E9",
                color = "white", 
                alpha = 0.9,
                bins = 25) +
  labs(x = "Score",
       y = "Number of Observations",
       title = "Distribution of Scores")
```


## Description of the remaining 54 variables 

```{r, include=FALSE}
str(data)
```

### ID Variables

| **Variable ID** |  **Variable Description**    | 
| :-------------- | :--------------------------: | 
| `id`            | Student identifier           |
| `attnd_dist_inst_id` | Institution identifier for the Attending District |
| `attnd_schl_inst_id` | Institution identifier for the Attending School |
| `partic_dist_inst_id` | Institution identifier for the Attending District |
| `partic_schl_inst_id` | Institution identifier for the Attending School |
| `ncessch` | School id number |

### Categorical Predictor Variables 

| **Variable ID** |  **Variable Description**    | 
| :-------------- | :--------------------------: | 
| `gndr`            | Student gender       |
| `ethnic_cd`         | Student ethnicity       |
| `enrl_grd`          | Student grade level       |
| `calc_admn_cd`      | Code describing special circumstances affecting test administration |
| `tst_bnch`          | Benchmark level of the administered test       |
| `tst_dt`            | Test date       |
| `migrant_ed_fg`     | Indicates student participation in a program designed to assure that migratory children receive full and appropriate opportunity to meet the state academic content and student academic achievement standards       |
| `ind_ed_fg`         | Indicates student participation in a program designed to meet the unique educational and culturally related academic needs of American Indians       |
| `sp_ed_fg`          | Indicates student participation in an Individualized Education Plan (IEP/IFSP)       |
| `tag_ed_fg`         | Indicates student participation in a Talented and Gifted program       |
| `econ_dsvntg`       | Indicates student eligibility for a Free or Reduced Lunch program       |
| `ayp_lep`           | Indicates a student who received services or was eligible to receive services in a Limited English Proficient program       |
| `stay_in_dist`            | Indicates that the student has been enrolled for more than 50% of the days in the school year as of the first school day in May at the district where the student is resident on the first school day in May       |
| `stay_in_schl`      | Indicates that the student has been enrolled for more than 50% of the days in the school year as of the first school day in May at the school where the student is resident on the first school day in May      |
| `dist_sped`         | Indicates that the student was enrolled in a district special education program during the school year and received general education classroom instruction for less than 40% of the time as of the first school day in May       |
| `trgt_assist_fg`    | Flag indicating the record is included in Title 1 Targeted Assistance for the Adequate Yearly Progress (AYP) school performance calculations       |
| `ayp_dist_partic`   | Flag indicating the record is included in the denominator of Adequate Yearly Progress (AYP) district participation calculations       |
| `ayp_schl_partic`   | Flag indicating the record is included in the denominator of Adequate Yearly Progress (AYP) school participation calculations       |
| `ayp_dist_prfrm`    | Flag indicating the record is included in the denominator of Adequate Yearly Progress (AYP) district performance calculations       |
| `ayp_schl_prfrm`    | Flag indicating the record is included in the denominator of Adequate Yearly Progress (AYP) school performance calculations       |
| `rc_dist_partic`    | Flag indicating the record is included in the denominator of Report Card (RC) district participation calculations       |
| `rc_schl_partic`    | Flag indicating the record is included in the denominator of Report Card (RC) school participation calculations       |
| `rc_dist_prfrm`     | Flag indicating the record is included in the denominator of Report Card (RC) district performance calculations       |
| `rc_schl_prfrm`     | Flag indicating the record is included in the denominator of Report Card (RC) school participation calculations       |
| `lang_cd`           | Test language       |
| `tst_atmpt_fg`      | Code describing whether the test was attempted       |
| `grp_rpt_dist_partic`  | Flag indicating the record is included in the denominator of Group Report district participation calculations       |
| `grp_rpt_schl_inst_id` | Flag indicating the record is included in the denominator of Group Report school participation calculations       |
| `grp_rpt_dist_prfrm`   | Flag indicating the record is included in the denominator of Group Report district performance calculations       |
| `grp_rpt_schl_prfrm`   | Flag indicating the record is included in the denominator of Group Report school participation calculations       |
| `sch_name`          | School name       |

#### Gender Distribution

```{r, include=FALSE}

gndr_count <- data %>% 
  count(gndr)

gndr_plot <- ggplot(gndr_count, aes(gndr, n)) +
  geom_col(fill = "#56B4E9",
                color = "white", 
                alpha = 0.9) +
  labs(x = "Gender",
       y = "Total")

```

```{r, include=TRUE, fig.height=8, fig.width=11}
gndr_plot
```

#### Ethnicity Distribution 

```{r, include=FALSE}
eth_count <- data %>% 
  count(ethnic_cd)

eth_plot <- ggplot(eth_count, aes(fct_reorder(ethnic_cd, n), n)) +
    geom_col(fill = "#56B4E9",
                color = "white", 
                alpha = 0.9) +
  coord_flip() +
  labs(x = "Ethnic Id",
       y = "Total",
       caption = "W = White race/non-Hispanic\n H = Hispanic ethnicity\n M = multi-racial, non-Hispanic\n A = Asian race, non-Hispanic\n B = Black/African American, non-Hispanic\n I = American Indian/Alaskan Native race, non-Hisapnic\n P = Pacific Islander, non-Hispanic")
```

```{r, include=TRUE, fig.height=8, fig.width=11}
eth_plot
```

### Continuous Predictor Variables 

| **Variable ID** |  **Variable Description**    | 
| :-------------- | :--------------------------: | 
| `lat`            | School latitude           |
| `lon` | school longitude  |
| `p_american_indian_alaska_native` | Percentage of students identified as American Indian and/or Alaska Native |
| `p_asian` | Percentage of students identified as Asian |
| `p_native_hawaiin_pacific_islander` | Percentage of students identified as Native Hawaiin and/or Pacific Islander |
| `p_black_african_american` | Percentage of students identified as black and/or African American |
| `p_hispanic_latino`            | Percentage of students identified as Hispanic and/or Latino           |
| `p_white` | Percentage of students identified as white |
| `p_multiracial` | Percentage of students identified as multiracial |
| `free_lunch_qualified` | Total number of students who qualify for free lunch |
| `reduced_price_lunch_qualified` | Total number of students who qualify for reduced price lunch |
| `missing` |  |
| `not_applicable`            |            |
| `no_category_codes` |  |
| `n` | Total number of students enrolled in the school |
| `prop_frl` | Proportion of students who qualify for free lunch |
| `prop_reduced_lunch` | Proportion of students who qualify for reduced price lunch |

#### Distribution of Ethnicities 

```{r, include=FALSE}

ggplot(data, aes(p_asian)) +
  geom_histogram(fill = "#56B4E9",
                color = "white", 
                alpha = 0.9,
                bins = 25) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Score",
       y = "Ethnicity Percentage",
       title = "Distribution of Scores")

percent_hist <- function(df, x) {
  ggplot(data, aes({{x}})) +
  geom_histogram(fill = "#56B4E9",
                color = "white", 
                alpha = 0.9,
                bins = 15) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Percentage",
       y = "Number of Schools")
}

percent_hist(data, p_asian) +
  labs(title = "Asian")

ethnic_data_long <- data %>% 
  pivot_longer(
    cols = c(41:47),
    names_to = "ethnicity",
    names_prefix = "p_",
    values_to = "percent"
  )
    
```

##### American Indian/Alaskan Native race, non-Hispanic

```{r, include=TRUE, fig.height=8, fig.width=11}
percent_hist(data, p_american_indian_alaska_native)
```

##### Asian, non-Hispanic

```{r, include=TRUE, fig.height=8, fig.width=11}
percent_hist(data, p_asian)
```

##### Black/African American, non-Hispanic

```{r, include=TRUE, fig.height=8, fig.width=11}
percent_hist(data, p_black_african_american)
```

##### Hispanic 

```{r, include=TRUE, fig.height=8, fig.width=11}
percent_hist(data, p_hispanic_latino)
```

##### Multi-racial, non-Hispanic 

```{r, include=TRUE, fig.height=8, fig.width=11}
percent_hist(data, p_multiracial)
```

##### Pacific Islander race, non-Hispanic

```{r, include=TRUE, fig.height=8, fig.width=11}
percent_hist(data, p_native_hawaiian_pacific_islander)
```

##### White 

```{r, include=TRUE, fig.height=8, fig.width=11}
percent_hist(data, p_white)
```

#### Distribution of Free and Reduced Lunch Status

Distribution of the proportion of students who qualify for free lunch 

```{r, include=TRUE, fig.height=8, fig.width=11}
ggplot(ethnic_data_long, aes(prop_frl)) +
  geom_histogram(fill = "#56B4E9",
                color = "white", 
                alpha = 0.9,
                bins = 15) +
  scale_x_continuous(labels = scales::percent) +
  facet_wrap(~ethnicity) +
  labs(x = "Percentage",
       y = "Number of Schools")
```

Distribution of the proportion of students who qualify for reduced price lunch 

```{r, include=TRUE, fig.height=8, fig.width=11}
ggplot(ethnic_data_long, aes(prop_reduced_lunch)) +
  geom_histogram(fill = "#56B4E9",
                color = "white", 
                alpha = 0.9,
                bins = 15) +
  scale_x_continuous(labels = scales::percent) +
  facet_wrap(~ethnicity) +
  labs(x = "Percentage",
       y = "Number of Schools")
```

```{r}
summary(data)
```

# Split and Resample 

```{r, include=TRUE}
set.seed(3000)
data_split <- initial_split(data, strata = "score")

set.seed(3000)
train <- training(data_split)
test <- testing(data_split)

set.seed(3000)
data_cv <- vfold_cv(train, strata = "score")

```

# Preprocess 

We processed the varibales in the dataset through the following steps: 

- All id variables are assigned to "id" role, which are not used as predictors  
- All nominal varibales, except the outcome, with a missing value in a factor level are assigned to "unknown" level. 
- All numeric variables with a missing value are imputed by its median value.
- Predictor variables with nearly no variation are removed
- All nominal varibales, except id variables and outcomes, are dummy coded
- After dummny coded, all nominal variables with nearly no variation are removed. 

After processing the data, we are left with 1 outcome variable, 7 id variables, and 39 predictors. 
```{r, include=TRUE}
# this from model 1
rec <- recipe(score ~ ., train) %>% 
  step_mutate(tst_dt = as.numeric(lubridate::mdy_hms(tst_dt))) %>% 
  update_role(contains("id"), ncessch, new_role = "id vars") %>% 
  step_novel(all_nominal()) %>%
  step_unknown(all_nominal()) %>%
  step_nzv(all_predictors()) %>%
  step_medianimpute(all_numeric(), -all_outcomes(), -has_role("id vars")) %>%
  step_dummy(all_nominal(), -has_role("id vars")) %>%
  step_nzv(all_predictors()) %>% 
  step_interact(~lat:lon)

prep(rec)  

baked_train <- rec %>% prep() %>% bake(train)
baked_train
```
