---
title: "Description of the dataset"
description: |
  Core features of the data, variable transformations, and data splitting process.
author:
  - name: Jim - Claire - Thuy
    url: https://www.uoregon.edu/
date: 12-04-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(ggplot2)
library(tidymodels)
```

# Description of the data 

The original dataset has 39 variables. We join this data with the "Fall membership report 2019 - 2020", which contains the information about students' ethnicities, adding 8 more variables to make a dataset of total 47 variables. 

```{r}
set.seed(3000)
data <- read_csv(here::here("data", "train.csv")) %>% 
  select(-classification)

data <- dplyr::sample_frac(data, size = 0.01)

sheets <- readxl::excel_sheets(here::here("data",
"fallmembershipreport_20192020.xlsx"))

ode_schools <- readxl::read_xlsx(here::here("data",
"fallmembershipreport_20192020.xlsx"), sheet = sheets[4])

ethnicities <- ode_schools %>%
select(attnd_schl_inst_id = `Attending School ID`,
sch_name = `School Name`,
contains("%")) %>%
janitor::clean_names()
names(ethnicities) <- gsub("x2019_20_percent", "p", names(ethnicities))

data <- left_join(data, ethnicities)

head(data)
```
We're interested in predicting the students' score based on their attributes. The range of the outcome variable is as follows:

```{r}
summary(data$score)
```

Discription of the predictive variables:

```{r}
summary(data)
```

# Split and Resample 

```{r, include=TRUE}
set.seed(3000)
data_split <- initial_split(data, strata = "score")

set.seed(3000)
train <- training(data_split)
test <- testing(data_split)

set.seed(3000)
data_cv <- vfold_cv(train, strata = "score")

```

# Preprocess 

We processed the varibales in the dataset through the following steps: 

- All id variables are assigned to "id" role, which are not used as predictors  
- All nominal varibales, except the outcome, with a missing value in a factor level are assigned to "unknown" level. 
- All numeric variables with a missing value are imputed by its median value.
- Predictor variables with nearly no variation are removed
- All nominal varibales, except id variables and outcomes, are dummy coded
- After dummny coded, all nominal variables with nearly no variation are removed. 

After processing the data, we are left with 1 outcome variable, 7 id variables, and 39 predictors. 
```{r, include=TRUE}
rec <- recipe(
    formula = score ~ ., data = train 
  ) %>%
 step_mutate(tst_dt = lubridate::mdy_hms(tst_dt)) %>%
 update_role(contains("id"), ncessch, sch_name, new_role = "id") %>%
 # step_novel(all_nominal(), -all_outcomes()) %>%
 step_unknown(all_nominal(), -all_outcomes()) %>%
 step_medianimpute(all_numeric()) %>%
 step_nzv(all_predictors(), freq_cut = 0, unique_cut = 0) %>%
 step_dummy(all_nominal(), -has_role(match = "id"), -all_outcomes()) %>%
 step_nzv(all_predictors())

prep(rec)  
```
